Neural Connectome of the Ctenophore Statocyst

### 2025-05-23 manuscript/Manuscript.qmd
MS: edits, add full connectome code, source_data folder

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973a]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).

@@

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) Morphological rendering of the large aboral nerve-net neuron ANN_Q1-4 spanning all four quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel). Spheres indicate the positions of nuclei.\
(B) Morphological rendering of the two smaller ANN neurons ANN_Q1Q2 (pink) and ANN_Q3Q4 (orange), each spanning two quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel).  Spheres indicate the positions of nuclei.\
(C) Number of mitochondria per cell for each cell type. The number of synapse-associated mitochondria is shown in green, the number of mitochondia outside synapses is shown in grey. N > 2 cells for each cell type except lithocytes. Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cells; imc, intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl, plumose; ef, epithelial floor cells.\
(D) Position of mitochondria within ANN_Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv, synaptic vesicles.\
(F) Position of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN_Q1-4.](figures/Figure2.png)

@@

ANNs form no synaptic output to classical effector organs such as muscles or comb plates. All observed output synapses were directed to balancer ciliary cells. Moreover, the projection patterns of individual ANNs were asymmetric, selectively targeting different regions of the balancer system. These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense. Rather, they may be involved in coordinating or modulating the timing and dynamics of the activity of the sensory-motile balancer cilia, possibly modulating their responsiveness to statocyst load [@keijzer2013; @dewiljes2015; @Veraszto2017].

What could be the function of the aboral nerve net? We found that the anatomical differences in the innervation patterns of ANNs across the sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This suggests that individual ANNs may play opposing or complementary roles in regulating ciliary beating---potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved [@Ryan2013;@Moroz2014]. However, the combination of circuit architecture and behavioural outputs provides the first functional insight into the role of neurons in a ctenophore.

Strikingly, this organization resembles circuit motifs found in other animals. For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024]. Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015; @roberts2022].

Taken together, our findings suggest that the ANN network represents a non-motor, coordination-based neural system—one whose primary function is to tune and synchronize locomotor effectors rather than activate them directly. This departs fundamentally from reflex-arc models of early nervous systems [@arendt2021; @jékely2010] and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons.

![**Figure 5. Comparison of Neural Circuits Regulating Ciliary Movement**\
(A) Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation. Ciliated cells are shown in gray. Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
(Left) Gravity-sensing neural circuit of *M. leidyi*, as revealed in this study. Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
(Righr) Neural circuit regulating prototroch ciliary movement in *P. dumerilii*. Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while cholinergic neurons (MC neurons, magenta) induce ciliary arrest. Arrows directed toward ciliated cells in the head prototroch are color-coded accordingly.](figures/Figure5.png)

@@

In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ. Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @tamm2014b]. Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024]. These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.

As revealed by our high-speed imaging, ciliary beat frequency (CBF) exhibited rhythmic changes over time that differed between anatomical planes (Figures 4E and F), indicating spatially distinct coordination dynamics. These CBF rhythms are likely generated by the ANN circuit, while bridge cells may function to stabilize or adjust this rhythmic output in accordance with the real-time state of the cilia. In this way, bridge cells could provide feedback about the execution target—i.e., the balancer cilia—back into the coordination circuit, serving as a core element of a self-regulating rhythm control system.

This architecture can be interpreted as a recurrent information loop, in which output from the ANN circuit modulates ciliary activity, and the resulting ciliary state is in turn relayed back to the ANN via bridge cells. Such a structure suggests a loop-based coordination mechanism [@arshavsky2003; @selverston2010], where the central circuit adapts its timing based on feedback from its target. Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may play a central role in maintaining the robustness and flexibility of endogenous ciliary rhythms [@kennedy2024].

@@

How do differences in balancer cilia coordination across anatomical planes shape behaviour? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020]. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.

Our findings reveal a circuit composed of ANN neurons and bridge cells that coordinates ciliary beating rhythms in space and time without sensory input, and features a recursive architecture in which ciliary outputs are relayed back into the circuit via bridge cells. Such a configuration suggests that even with a limited number of cells, the system is capable of flexible and robust rhythm control.

As Jagar et al. (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011]. The network we describe here similarly orchestrates output through dynamic, self-regulating coordination. These dynamics, visualized through live imaging of the entire cell ensemble, offer a window into the core functions of a nervous system—not in complexity, but in principle.

@@

![**Figure 1—figure supplement 1**\
(A)...](figures/Figure1_Supplement1.png)

>>>>

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a]. In our dataset, we identified 428 synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Figure 2--figure supplement 1A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973a]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Figure 2--figure supplement 1B).

@@

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) Morphological rendering of the large aboral nerve-net neuron ANN_Q1-4 spanning all four quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel). Spheres indicate the positions of nuclei.\
(B) Morphological rendering of the two smaller ANN neurons ANN_Q1Q2 (pink) and ANN_Q3Q4 (orange), each spanning two quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel). Spheres indicate the positions of nuclei.\
(C) Number of mitochondria per cell for each cell type. The number of synapse-associated mitochondria is shown in green, the number of mitochondia outside synapses is shown in grey. N \> 2 cells for each cell type except lithocytes. Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cells; imc, intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl, plumose; ef, epithelial floor cells.\
(D) Position of mitochondria within ANN_Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv, synaptic vesicles.\
(F) Position of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN_Q1-4.](figures/Figure2.png)

@@

ANNs form no synaptic output to classical effector organs such as muscles or comb plates. All observed output synapses were directed to balancer ciliary cells, bridge cells or other sensory cells. This suggests that ANNs are unlikely to serve as motor drivers in the conventional sense. Rather, they may be involved in coordinating or modulating the timing and dynamics of the activity of the sensory-motile balancer cilia, possibly modulating their responsiveness to statocyst load [@keijzer2013; @dewiljes2015; @Veraszto2017].

What could be the function of the aboral nerve net? We found differences in the innervation patterns of ANNs across the sagittal and tentacular planes. The two smaller ANNs targeted a pair of balancers (either Q1-Q2 or Q3-Q4) across the sagittal plane. This innervation pattern corresponded with plane-specific differences in the dynamics of balancer cilia. Across the sagittal but not the tentacular plane, the closures of balancer cilia were near-simultaneous, within less than one beat phase. In contrast, the large ANN innervated balancers in all four quadrants and could account for the near-simultaneous re-beat (within less than one beat phase) of the four balancers. The neurotransmitter identity and activation dynamics of these neurons remains unresolved. However, the combination of circuit architecture and behavioural outputs provides the first functional insight into the role of neurons in a ctenophore.

The functional organisation of the ANN circuit is reminiscent of the ciliomotor circuit in the larvae of the marine annelid *Platynereis dumerilii*. Here, the activation of cholinergic motor neurons leads to the synchronous arrest of cilia on multiple multiciliary bands [@Veraszto2017]. This is antagonised by serotonergic midline-crossing neurons that drive ciliary re-beat [@Veraszto2017; @calderon2024]. The ctenophore balancer-circuit is the second synaptically mapped ciliomotor circuit that convergently evolved analogous control mechanisms to the annelid and other cilimotor systems [@moroz2015; @Marinkovi2020; @roberts2022].

Overall, our findings suggest that the ANN network has a coordinating function to synchronize the activity of balancer cilia thereby possibly fine-tuning their activity or resetting their sensitivity. This departs fundamentally from reflex-arc models of simple nervous systems [@arendt2021; @jekely2010] and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons [@keijzer2013]. Such an organisation may have been retained in modern ctenophores ever since they evolved a pelagic lifestyle and a balancer [@stanley1983].

![**Figure 5. Comparison of Neural Circuits Regulating Ciliary Movement**\
(A) Neurons are represented by circles, color-coded to indicate analogous functions in ciliary control. Ciliated cells are shown in gray. Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
(Left) Neural circuit of the *M. leidyi* gravi-sensory organ. Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying a potential involvement in feedback mechanisms between neurons and ciliated cells.\
(Right) Neural circuit regulating prototroch ciliary movement in *P. dumerilii*. Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while a cholinergic neuron (MC neuron, magenta) induces ciliary arrest.](figures/Figure5.png)

@@

In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ. Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections towards the base of the balancer cilia [@Tamm_2002; @tamm2014b]. Based on this spatial relationship, they may receive information about the physical state of the cilia---such as mechanical load or phase---or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024]. These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed towards the ANN circuit.

As revealed by our high-speed imaging, ciliary beat frequency (CBF) exhibited rhythmic changes over time that differed between anatomical planes (Figures 4E and F), indicating spatially distinct coordination dynamics. These CBF rhythms are likely generated by the ANN circuit, while bridge cells may function to stabilize or adjust this rhythmic output in accordance with the real-time state of the cilia. This way, bridge cells may provide feedback about the state of their cilia to the coordination circuit, serving as a core element of a self-regulating rhythm-control system.

This architecture can be interpreted as a recurrent information loop, in which output from the ANN circuit modulates ciliary activity, and the resulting ciliary state is in turn relayed back to the ANN via bridge cells. Such a structure suggests a loop-based coordination mechanism [@arshavsky2003; @selverston2010], where the central circuit adapts its timing based on feedback from its target. Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may function in maintaining the robustness and flexibility of endogenous ciliary rhythms [@kennedy2024].

@@

How do differences in balancer cilia coordination across anatomical planes shape behaviour? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020]. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015].

Our findings reveal a circuit composed of ANN neurons and bridge cells that coordinates ciliary beating rhythms in space and time without sensory input, and features a recursive architecture in which ciliary outputs may feed back to the circuit via bridge cells. Such a configuration suggests that even with a limited number of cells, the system is capable of flexible and robust rhythm control.

As Jagar et al. (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011]. The network we describe here orchestrates output through dynamic, self-regulating coordination. Future work could uncover this dynamics, visualized through live imaging of the entire cell ensemble, to offer a window into the core functions of a unique nervous system.

@@

![**Figure 1—figure supplement 1. Cell-type composition of the ctenophore aboral organ**\
Morphological rendering of skeletonised cell types, (A) balancer cells (B) bridge cells, (C) bristle cells, (D) ciliated groove cells, (E) dense vesicle cells, (F) dome cells, (G) intra-multiciliated cells, (H) lamellate bodies, (I) lithocytes, (J) plumose cells, (K) aboral nerve-net neurons (L) epithelial floor cells.](figures/Figure1_Supplement1.png)

![**Figure 1—figure supplement 2. Balancers with all other cells in the ctenophore aboral organ**\
Morphological rendering of skeletonised cells with the balancer cells shown in orange.](figures/Figure1_Supplement2.png)
